---
title: "Current Status of the Implementation:\\newline Phylogenetic Peeling Algorithm (Project 2)"
author: Cores D+C
date: February 23, 2017
fig_width: 8
fig_height: 6
fontsize: 10pt
output: beamer_presentation
header-includes:
  - \usepackage{hyperref}
  - \hypersetup{allcolors=blue, colorlinks=true}
---

## Table of Contents

\tableofcontents

# Phylogenetics R package

## Phylogenetics R package

*   Implements phylogenetic peeling algorithm.

*   The model has parameters:
    
    *   $\pi$ Root node probabilities.
    *   $\psi$ Misclassification probabilities.
    *   $\mu$ Gain/Loss probabilities.
    
    The log-likelihood function is written completely in C++ (RcppArmadillo)

*   Currently 3 different methods for estimating the model: Newton-Raphson
    and Artificial Bee Colony Optimization (ABC) in the side of MLE, and MCMC. **(all three allow using priors)**
    
*   Besides, has a bunch of functions for processing and visualizing data.

*   Let's see an example!

-----

### Phylogenetic Example

\footnotesize

```{r phylo-data}
library(aphylo)

# Loading and taking a look at the data
data("experiment")
head(experiment)
data("tree")
head(tree)
```

\normalsize

-----

### Phylogenetic Example (cont. 1)

\footnotesize

*   Make the data suitable for the peeling algorithm.

*   Also, we are borrowing methods from the `ape` package (like the plot).

```{r phylo-offspring-data, out.width=".5\\linewidth"}
# Preparing the data 
O <- get_offspring(experiment, "LeafId", tree, "NodeId", "ParentId")
plot(O)
```

\normalsize

-----

### Phylogenetic Example (cont. 2)

*   We also have our own methods

\footnotesize

```{r phylo-offspring-loglike, out.width=".7\\linewidth", echo=FALSE, fig.cap="The `plot\\_LogLike` function", cache=TRUE}
# Nice personalized plot
plot_LogLike(O, nlevels = 60, plotfun = persp, theta = -pi*20, 
  shade=.7, border="darkblue", phi=30, scale=TRUE, 
  par.args = list(mar=c(1, 1, 1, 1), oma=c(0,0,4,0)),
  ticktype = "detailed")
  
# Adding title
mtext(
  "LogLikelihood",
  side=3, outer=FALSE, line = 2, cex=1.25)
```

\normalsize

-----

### Phylogenetic Example (cont. 3)

*   Here is an example using the `phylo_mle`

\footnotesize

```{r mle, cache=TRUE}
# Computing Estimating the parameters 
ans_nr  <- phylo_mle(rep(.5,5), dat=O)
ans_abc <- phylo_mle(rep(.5,5), dat=O, useABC = TRUE,
                     control = list(maxCycle = 50))
```

```{r mle-plot, echo=FALSE, fig.width=8, fig.height=3.5, out.width='.8\\linewidth'}
oldpar <- par(no.readonly = TRUE)
par(mfrow=c(1,2))
plot(ans_nr, main = "MLE using Newton-Raphson")
plot(ans_abc, main = "MLE using ABCoptim")
par(oldpar)
```

\normalsize

-----

### Phylogenetic Example (cont. 4)

*   Using priors is very straight forward, e.g. $\psi\sim Beta(2,10)$

\footnotesize

```{r mle-prior, cache=TRUE}
ans_w_prior <- phylo_mle(rep(.5,5), dat=O, useABC = TRUE,
                     priors = function(pars) dbeta(pars[1:2], 2, 10),
                     control = list(maxCycle = 50))
```

```{r mle-prior-plot, echo=FALSE,fig.width=4, fig.height=4, out.width='.5\\linewidth'}
plot(ans_w_prior, main = "MLE using Newton-Raphson\n With Beta(2,10) priors")
```

\normalsize

## R Package to-do list

*   Change the name

*   Stare at the Newton-Raphson implementation (still returns funny numbers...)

*   Implement (or grab) a phylogenetic tree simulation routine (`ape` and `igraph` have some candidates)

*   Run more test



# MCMC function

## MCMC

*   The `MCMC` function implements the Metropolis-Hastings algorithm

*   It uses a symmetric transition kernel: A normal random-walk with scale
    parameter $s$:
    
    $$
    \theta' = \theta + s\times Z,\quad Z\sim N(0,1)
    $$

*   Furthermore, using reflecting boundaries, it allows specifying upper
    and lower bounds $\theta' \in \left[\underline \theta, \overline\theta\right]$
    
*   Just like the `mcmc::metrop` function, the user needs to pass a function that
    computes a the log unnormalized probability density.
    
*   Preliminary tests show that `MCMC` is 3x faster than `mcmc::metrop` (!).
    
*   Lets see an example!

-----

### MCMC example

*   Here we are simulating 1,000 data points $x_i\sim N(2.6, 3)$

*   We are also defining the log unnormalized density function, `fun`.
    Notice that the only parameter required in `fun` is the vector of
    parameters itself (data `D` is passed by R's scoping rules).

\footnotesize

```{r MCMC sim, size='small'}
# Loading the package
library(aphylo)

# Simulating data
set.seed(1231)
n    <- 1e3
pars <- c(mean = 2.6, sd = 3)

# Generating data and writing the log likelihood function
D   <- rnorm(n, pars[1], pars[2])
fun <- function(pars) sum(log(dnorm(D, pars[1], pars[2])))
```

\normalsize

-----

### MCMC example (cont. 1)

*   This call of the function has the (current) complete list of arguments
    that can be used in it.

\footnotesize

```{r MCMC call, cache=TRUE}
system.time(
ans <<- MCMC(
  fun     = fun,              # The log unormalized dens
  initial = c(mu=1, sigma=1), # Initial state (names are optional)
  nbatch  = 1e5,              # Number of steps
  burnin  = 1e3,              # Burn-in
  thin    = 20,               # Thinning
  scale   = .1,               # Scale of transition kernel
  ub      = 10,               # Upper bound (same for both params)
  lb      = c(-10, .5),       # Lower bound
  useCpp  = TRUE              # Request using the C++ version
  ))
```

\normalsize

-----

### MCMC example (cont. 2)

*   `ans`, is an object of class `mcmc` from the `coda` package.

```{r MCMC-plot, echo=FALSE, fig.cap="Calling of `plot(ans)` from the coda package.", out.width=".8\\linewidth", cache=FALSE}
library(coda)
plot(ans)
```


-----

### MCMC example (cont. 3)


```{r MCMC-autocorrplot, echo=FALSE, fig.cap="Calling of the `autocorr.plot(ans)` from the coda package.", out.width=".8\\linewidth", cache=FALSE}
autocorr.plot(ans)
```

## MCMC to-do list

*   Tempering MCMC.

*   Adaptative MCMC (global and local).

*   Parallel MCMC (multiple chains)... although this is easy to do with the `parallel` package

*   (another) Just Another Gibbs Sampler? Maybe RcppJAGS? (take a look at
    rcppbugs [here](https://github.com/armstrtw/rcppbugs) and [here](http://past.rinfinance.com/agenda/2012/talk/WhitArmstrong.pdf))

*   ...

\footnotesize

```cpp
// Reflexion adjustment
for (int k=0; k<K; k++) {
  
  while( (ans[k] > ub[k]) | (ans[k] < lb[k]) ) {
    
    if (ans[k] > ub[k]) {
      ans[k] = 2.0*ub[k] - ans[k];
    } else {
      ans[k] = 2.0*lb[k] - ans[k];
    }  
    
  }
  
}
```

\normalsize
